Geddy Lucier
Merging Data Script
Scuba.Tech Candlewood Lake


```{r}
library(tidyverse)
library(sf)
library(matrixStats)  # For doing fast rowwise operations
library(terra)
library(readxl)
library(ggplot2)
library(ggforce)
library(stringr)
library(cowplot)
library(dplyr)

setwd("Dropbox/Scuba.Tech/SdT-CT_AIS-Geddy-Winter-2024")
getwd()


```



```{r}

# Read the first file as the base dataset
merged_transect <- read_csv(file_list[1])

# Standardize column names (optional)
clean_names <- function(df) {
  colnames(df) <- tolower(trimws(gsub("[^A-Za-z0-9]", "_", colnames(df))))
  return(df)
}

merged_transect <- clean_names(merged_transect)

# Iterate over the rest of the files
for (i in seq_along(file_list)[-1]) {  # Skip first file since it's already loaded
  temp_transect <- read_csv(file_list[i])
  temp_transect <- clean_names(temp_transect)

  # Ensure both data frames have the same column structure
  all_cols <- union(colnames(merged_transect), colnames(temp_transect))

  # Add missing columns as NA in both data frames
  for (col in setdiff(all_cols, colnames(merged_transect))) {
    merged_transect[[col]] <- NA
  }
  for (col in setdiff(all_cols, colnames(temp_transect))) {
    temp_transect[[col]] <- NA
  }

  # Ensure column order matches
  merged_transect <- merged_transect[, all_cols]
  temp_transect <- temp_transect[, all_cols]

  # Append data
  merged_transect <- rbind(merged_transect, temp_transect)
  message(paste("Appended data from", file_list[i]))
}

# View the result
glimpse(merged_transect)



```
For final troubleshoots with OCR cleaning

```{r}
# elasp elesp are og
getOption("max.print" = 1000) 



merged_transect$date[merged_transect$date >=1] 

options(max.print=1000000)
unique(merged_transect)




```


```{r}
merged_transect <- read_csv("merged_transect.csv") #START HERE

merged_transect <- merged_transect %>%
  arrange(transect, point)

merged_transect <- merged_transect[-35]



#merged_transect$date[merged_transect$year == "0.371215278" ]

merged_transect <- merged_transect %>%
  mutate(year = gsub("(.*)/([0-9]+)$", "20\\2", date),
          myrspi = ifelse(myrspi < 0 | myrspi > 5, 0, myrspi)
         ) %>%
  group_by(year) %>%
  mutate(myspri.yearsum = sum(myrspi),
         cerdem.yearsum = sum(cerdem)) # to create year variable


unique(merged_transect$year)
ggplot() +  geom_line(aes(data = merged_transect, x = year, y =))

write_csv(merged_transect, "merged_transect.csv")


```
temporary troubleshooting - disregard for later

```{r}
substrate <- merged_transect %>%
  filter(year != 2007) %>%
  select(substrate, year ,dfs_m_, depth_m_)



```



percentage of total

```{r}
?across

library(dplyr)

species_data <- merged_transect %>%
  relocate(year, .before = 1) %>%  # Move 'year' to the first column
  relocate(35, .after = year) %>% 
  dplyr::select(transect, point, lattitude, longitude, year, 13:ncol(.))  %>% # Move column 35 next to 'year'
  mutate(point_sum = rowSums(across(6:ncol(.)), na.rm = TRUE), ) %>% #sum of all the species within a point 
  group_by(year) %>% # Sum across columns 12 to end
  mutate(yearly_sum = sum(point_sum)) %>%
  mutate(potfol_sum = sum(potfol, na.rm = TRUE), 
         elonut_sum = sum(elonut, na.rm = TRUE),
         cerdem_sum = sum(cerdem, na.rm = TRUE),
         najfle_sum = sum(najfle, na.rm = TRUE),
         potcri_sum = sum(potcri, na.rm = TRUE),
         myrspi_sum = sum(myrspi, na.rm = TRUE),
         valame_sum = sum(valame, na.rm = TRUE),
         najmin_sum = sum(najmin, na.rm = TRUE),
         potgra_sum = sum(potgra, na.rm = TRUE),
         potper_sum = sum(potper, na.rm = TRUE),
         zanpal_sum = sum(zanpal, na.rm = TRUE),
         calsp_sum = sum(calsp, na.rm = TRUE),
         nymodo_sum = sum(nymodo, na.rm = TRUE),
         lemmin_sum = sum(lemmin, na.rm = TRUE),
         spipol_sum = sum(spipol, na.rm = TRUE),
         stupec_sum = sum(stupec, na.rm = TRUE),
         elasp_sum = sum(elasp, na.rm = TRUE),
         potbic_sum = sum(potbic, na.rm = TRUE),
         potpus_sum = sum(potpus, na.rm = TRUE),
         unidentif3_sum = sum(unidentif3, na.rm = TRUE),
         elesp_sum = sum(elesp, na.rm = TRUE),
         spaspp_sum = sum(spaspp, na.rm = TRUE),
         typha_sum = sum(typha, na.rm = TRUE)) %>%
  filter(!year == "2007") %>%#filter to year to control for BAD data
    ungroup()

colnames(species_data)ƒ
head(species_data)
tail(species_data)
unique(merged_transect$typha)
colnames(merged_transect)


#### og code
other_data <- species_data %>%
  arrange(transect, point) %>%
  dplyr::select(-c(   "potfol",     "elonut",    "cerdem",     "najfle",     "potcri",     "myrspi",     "valame",     "najmin",  "potgra",     "potper",     "zanpal",     "calsp",      "nymodo",     "lemmin",  "spipol",     "stupec",     "elasp" ,     "potbic",     "potpus",     "unidentif3",  "elesp",      "spaspp",     "typha")) %>%
  dplyr::select(-cerdem_sum, -myrspi_sum) %>%
mutate(
    other_sum = rowSums(across(8:28))
  ) %>%
  dplyr::select(other_sum)





species_data <- cbind(species_data, other_data)

write_csv(species_data, "species_data.csv")


```


```{r}

ggplot(data = species_data) + 
  # Opaque area for "Total" sum
  geom_area(aes(x = year, y = yearly_sum, fill = "Total"), alpha = 0.5, show.legend = F) +  # Use fill instead of color
  
  # Line for "Total" sum
  geom_line(aes(x = year, y = yearly_sum, color = "Total"), linewidth = 1) +  # Use color, not fill
  
  # Myrspi points and lines
  geom_point(aes(x = year, y = myrspi_sum, color = "Eurasian Watermilfoil"), size = 1) + 
  geom_line(aes(x = year, y = myrspi_sum, color = "Eurasian Watermilfoil"), linewidth = 1) + 
  
  # Cerdem points and lines
  geom_point(aes(x = year, y = cerdem_sum, color = "Coontail"), size = 1) + 
  geom_line(aes(x = year, y = cerdem_sum, color = "Coontail"), linewidth = 1) + 
  
  # Other species points and lines
  geom_point(aes(x = year, y = other_sum, color = "Other"), size = 1) + 
  geom_line(aes(x = year, y = other_sum, color = "Other"), linewidth = 1) + 
  
  # Define color and fill palettes
  scale_color_manual(values = c("Eurasian Watermilfoil" = "#E74C3C", "Coontail" = "#3498DB", "Other" = "#27AE60", "Total" = "Grey")) +
  scale_fill_manual(values = c("Total" = "#95A5A6")) +  # Opaque gray for total area
  
  # Labels and theme improvements
  labs(color = "Species", fill = "Total Area", x = "Year", y = "Observed Sum of Observations",
       title = "Candlewood Lake, CT Invasive Species Populations") +
  theme_minimal(base_size = 12) +
  theme(
    panel.background = element_blank(),  
    plot.background = element_blank(),  
    legend.position = "top"
  ) +
  scale_x_continuous(
    breaks = seq(2007.5, 2020, by = 2.5),  # Adjust tick intervals (every 5 years)
    labels = c("Jul. 2008", "Aug. 2010", "Aug. 2012", "Aug. 2015", "Aug. 2017", "Aug. 2020")  # Custom labels
  ) 







```
Plotting (LET'S GET THIS SHIT DONE)

```{r}
lake_data <- st_read("shapefiles/Lake Polygon/NLA_2017_Lake_Polygon.dbf")

gdb_path <- "shapefiles/fef0ad6f-84a0-42e6-942a-ed3172551e0c.gdb"
layers <- st_layers(gdb_path)
ct_lakes <- st_read(gdb_path)



candlewood_bbox <- st_as_sfc(st_bbox(c(
  xmin = -73.52,  # Minimum longitude (left)
  xmax = -73.42,  # Maximum longitude (right)
  ymin = 41.43,   # Minimum latitude (bottom)
  ymax = 41.53    # Maximum latitude (top)
), crs = 4326))


ct_lakes <- st_transform(ct_lakes, crs = 4326)


candlewood_lake <- ct_lakes %>%
  filter(st_intersects(SHAPE, candlewood_bbox, sparse = FALSE)) %>%
  filter(!WBNAME == "Squantz Pond" & SHAPE_Length > 5100)

fivenum(candlewood_lake$SHAPE_Length)

##just a test plot 
ggplot() +
  geom_sf(data = candlewood_lake, fill = "lightblue", color = "darkblue") +
  labs(
    title = "Candlewood Lake Base Layer",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal()



```
```{r}
ggplot()

coord_gendata <-species_data

transect_heads <-  species_data %>%
    filter(point == 1 & lattitude > 41.4) ## lat and long coordinates


sf_points <- st_as_sf(transect_heads, coords = c("longitude", "lattitude"), crs = 4326) # geometry






## test plot with points now
ggplot() +
  geom_sf(data = candlewood_lake, fill = "lightblue", color = "darkblue") +
  labs(
    title = "Candlewood Lake Base Layer",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() + 
  geom_sf(data= sf_points)


```

Notes: two ways we can plot
We need some way to aggregate the points so it doesn't look too messy. 
```{r}
install.packages("scatterpie")  # Install the package
library(scatterpie)  # Load the package

transect_heads %>%
  filter(year == "2008") %>%
  ggplot() + 
  geom_sf(data = candlewood_lake, fill = "lightblue", color = "darkblue") +  # Base layer
  geom_scatterpie(aes(x = longitude, y = lattitude, group = point, 
                      r = sqrt(yearly_sum) / 2500),  # 🔹 Scale pie size dynamically
                  cols = c("myrspi_sum", "cerdem_sum", "other_sum")) +
   # ✅ Fix: Ulog()coord_sf() +  # ✅ Fix: Uyearly_sumcoord_sf() +  # ✅ Fix: Use coord_sf() instead of coord_equal()
  theme_void()   # Removes grid, axes, background



plot(eight_plot)

```
```{r}
library(ggplot2)
library(scatterpie)
library(sf)
library(cowplot)

# Define bounding boxes for North & South
north_bbox <- c(xmin = -73.52, xmax = -73.42, ymin = 41.48, ymax = 41.53)  # Upper half
south_bbox <- c(xmin = -73.52, xmax = -73.42, ymin = 41.43, ymax = 41.48)  # Lower half

# Define a custom color palette
species_colors <- c("myrspi_sum" = "#66c2a5", "cerdem_sum" = "#fc8d62", "other_sum" = "#8da0cb")

# North section plot
north_plot <- transect_heads %>%
  filter(year == "2008") %>%
  ggplot() + 
  geom_sf(data = candlewood_lake, fill = "lightblue", color = "black") +  
  geom_scatterpie(aes(x = longitude, y = lattitude, group = point, 
                      r = sqrt(yearly_sum) / 2500), 
                  cols = c("myrspi_sum", "cerdem_sum", "other_sum"),
                  color = "black", alpha = 0.8) +  
  scale_fill_manual(values = species_colors, name = "Species") +  # Custom colors
  coord_sf(xlim = c(north_bbox["xmin"], north_bbox["xmax"]), 
           ylim = c(north_bbox["ymin"], north_bbox["ymax"])) +  
  theme_void() +  
  theme(legend.position = "none") +  
  labs(title = "Northern Section of Candlewood Lake")

# South section plot
south_plot <- transect_heads %>%
  filter(year == "2008") %>%
  ggplot() + 
  geom_sf(data = candlewood_lake, fill = "lightblue", color = "black") +  
  geom_scatterpie(aes(x = longitude, y = lattitude, group = point, 
                      r = sqrt(yearly_sum) / 2500), 
                  cols = c("myrspi_sum", "cerdem_sum", "other_sum"),
                  color = "black", alpha = 0.8) +  
  scale_fill_manual(values = species_colors, name = "Species") +
  coord_sf(xlim = c(south_bbox["xmin"], south_bbox["xmax"]), 
           ylim = c(south_bbox["ymin"], south_bbox["ymax"])) +  
  theme_void() +  
  theme(legend.position = "none") +  
  labs(title = "Southern Section of Candlewood Lake")

species_colors <- c("myrspi_sum" = "#66c2a5", "cerdem_sum" = "#fc8d62", "other_sum" = "#8da0cb")

# Create a dummy scatterpie plot just to extract the legend
# Generate a dummy plot with a legend
legend_plot <- ggplot(data.frame(x = 0, y = 0, myrspi_sum = 1, cerdem_sum = 1, other_sum = 1)) +
  geom_scatterpie(aes(x = x, y = y, group = 1), 
                  cols = c("myrspi_sum", "cerdem_sum", "other_sum")) +
  scale_fill_manual(values = species_colors, name = "Species",
                    labels = c("Eurasian Watermilfoil", "Coontail", "Other Species")) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.size = unit(1, "cm"),  
    legend.text = element_text(size = 12),  
    legend.title = element_text(size = 14, face = "bold")
  )

# Extract the legend correctly as a grob
legend_grob <- ggplotGrob(legend_plot)
legend_grob <- legend_grob$grobs[which(sapply(legend_grob$grobs, function(x) x$name) == "guide-box")][[1]]

# Convert legend_grob into a valid grob for cowplot
legend_cowplot <- ggdraw() + draw_grob(legend_grob)


# Create title and subtitle
title_plot <- ggdraw() + 
  draw_label("Candlewood Lake Survey - 2008", fontface = "bold", size = 20, hjust = 0.5) +
  draw_label("Species Distribution by Transect", fontface = "italic", size = 14, hjust = 0.5, y = 0.8)

# Combine the legend into a cowplot-compatible object
legend_cowplot <- ggdraw() + draw_grob(legend_grob)

# Merge everything together
final_plot <- plot_grid(
  title_plot,               # Title & Subtitle
  legend_cowplot,           # Corrected legend
  plot_grid(north_plot, south_plot, ncol = 2),  # North & South Plots
  ncol = 1, 
  rel_heights = c(0.15, 0.05, 0.8),  # Adjusts proportions
  align = 'v',  # Aligns the components vertically
  axis = "lr"   # Aligns along the left and right axis
)

# Display the final plot

final_plot
# Display the final plot
ggsave("eight_plot.png", plot = final_plot, bg = "white")
?ggsave



```
```{r}

# Define bounding boxes for North & South
north_bbox <- c(xmin = -73.52, xmax = -73.42, ymin = 41.48, ymax = 41.53)  # Upper half
south_bbox <- c(xmin = -73.52, xmax = -73.42, ymin = 41.43, ymax = 41.48)  # Lower half

# Define a custom color palette
species_colors <- c("myrspi_sum" = "#66c2a5", "cerdem_sum" = "#fc8d62", "other_sum" = "#8da0cb")

# North section plot
north_plot <- transect_heads %>%
  filter(year == "2020") %>%
  ggplot() + 
  geom_sf(data = candlewood_lake, fill = "lightblue", color = "black") +  
  geom_scatterpie(aes(x = longitude, y = lattitude, group = point, 
                      r = sqrt(yearly_sum) / 2500), 
                  cols = c("myrspi_sum", "cerdem_sum", "other_sum"),
                  color = "black", alpha = 0.8) +  
  scale_fill_manual(values = species_colors, name = "Species") +  # Custom colors
  coord_sf(xlim = c(north_bbox["xmin"], north_bbox["xmax"]), 
           ylim = c(north_bbox["ymin"], north_bbox["ymax"])) +  
  theme_void() +  
  theme(legend.position = "none") +  
  labs(title = "Northern Section of Candlewood Lake")

# South section plot
south_plot <- transect_heads %>%
  filter(year == "2008") %>%
  ggplot() + 
  geom_sf(data = candlewood_lake, fill = "lightblue", color = "black") +  
  geom_scatterpie(aes(x = longitude, y = lattitude, group = point, 
                      r = sqrt(yearly_sum) / 2500), 
                  cols = c("myrspi_sum", "cerdem_sum", "other_sum"),
                  color = "black", alpha = 0.8) +  
  scale_fill_manual(values = species_colors, name = "Species") +
  coord_sf(xlim = c(south_bbox["xmin"], south_bbox["xmax"]), 
           ylim = c(south_bbox["ymin"], south_bbox["ymax"])) +  
  theme_void() +  
  theme(legend.position = "none") +  
  labs(title = "Southern Section of Candlewood Lake")

species_colors <- c("myrspi_sum" = "#66c2a5", "cerdem_sum" = "#fc8d62", "other_sum" = "#8da0cb")

# Create a dummy scatterpie plot just to extract the legend
# Generate a dummy plot with a legend
legend_plot <- ggplot(data.frame(x = 0, y = 0, myrspi_sum = 1, cerdem_sum = 1, other_sum = 1)) +
  geom_scatterpie(aes(x = x, y = y, group = 1), 
                  cols = c("myrspi_sum", "cerdem_sum", "other_sum")) +
  scale_fill_manual(values = species_colors, name = "Species",
                    labels = c("Eurasian Watermilfoil", "Coontail", "Other Species")) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.size = unit(1, "cm"),  
    legend.text = element_text(size = 12),  
    legend.title = element_text(size = 14, face = "bold")
  )

# Extract the legend correctly as a grob
legend_grob <- ggplotGrob(legend_plot)
legend_grob <- legend_grob$grobs[which(sapply(legend_grob$grobs, function(x) x$name) == "guide-box")][[1]]

# Convert legend_grob into a valid grob for cowplot
legend_cowplot <- ggdraw() + draw_grob(legend_grob)


# Create title and subtitle
title_plot <- ggdraw() + 
  draw_label("Candlewood Lake Survey - 2020", fontface = "bold", size = 20, hjust = 0.5) +
  draw_label("Species Distribution by Transect", fontface = "italic", size = 14, hjust = 0.5, y = 0.8)

# Combine the legend into a cowplot-compatible object
legend_cowplot <- ggdraw() + draw_grob(legend_grob)

# Merge everything together
final_plot <- plot_grid(
  title_plot,               # Title & Subtitle
  legend_cowplot,           # Corrected legend
  plot_grid(north_plot, south_plot, ncol = 2),  # North & South Plots
  ncol = 1, 
  rel_heights = c(0.15, 0.05, 0.8),  # Adjusts proportions
  align = 'v',  # Aligns the components vertically
  axis = "lr", # Aligns along the left and right axis
rel_widths = c(1, 0.7, 1)
  )
# Display the final plot

final_plot
# Display the final plot
ggsave("twenty_plot.png", plot = final_plot,bg = "white")
?ggsave



```








Regression

```{r}
library(glmnet)
install.packages("glmnet")
species_data <- na.omit(species_data$)  # Remove rows with any NA values


x <- model.matrix(myrspi_sum ~ ., data = species_data)[, -1]  # Remove intercept column
y <- species_data$myrspi_sum


set.seed(42)  # Reproducibility
train <- sample(1:nrow(x), nrow(x) * 0.7)  
test <- setdiff(1:nrow(x), train)

grid <- 10^seq(10, -2, length = 100)  # 100 lambda values from 10^10 to 10^-2 - check what this means 

# Fitting Ridge Regression Model - alpha of 0 means ridge 
ridge.mod <- glmnet(x[train,], y[train], alpha = 0, lambda = grid, thresh = 1e-6)

cv.ridge <- cv.glmnet(x[train,], y[train], alpha = 0)
best_lambda <- cv.ridge$lambda.min

ridge.final <- glmnet(x[train,], y[train], alpha = 0, lambda = best_lambda)
ridge.final

ridge.pred <- predict(ridge.final, s = best_lambda, newx = x[test,])
ridge.pred
mse <- mean((ridge.pred - y[test])^2)
print(mse)




```
Basemap Testing

```{r}
devtools::install_github("16EAGLE/basemaps")
library(basemaps)

get_maptypes()
set_defaults(map_service = "osm", map_type = "topographic")
basemap_test   <- basemap_magick(ext)

library(sf)


library(sf)

# Define bounding box in WGS84
ext <- st_as_sfc(st_bbox(c(xmin = -73.50, xmax = -73.40, ymin = 41.41, ymax = 41.55), crs = 3857))



# Now fetch the basemap
basemap_test <- basemap_gglayer(ext, map_service = "carto", map_type = "light_no_labels")



basemap_test
ggplot(basemap_test)

library(ggplot2)

# Create a ggplot and add the basemap layer
ggplot() +
  basemap_test +  # Add basemap_gglayer() output
  coord_sf() +  # Maintain correct spatial reference
  theme_void()  # Clean map with no axes or gridlines

```

Zebra Muscle incorporation
```{r}

zebra <- st_read("zeb_msc.kml")

plot(zebra)


```

Testing Phase for the jitter function
```{r}
species_data


geo_stats <- merged_transect %>%
  filter(year != "2007") %>%
  select(lattitude, longitude, dfs_m_, depth_m_)


speciesDT <- cbind(species_data, geo_stats)

set.seed(123)


speciesDT <- speciesDT[rep(1:nrow(speciesDT), each = 4), ] # Make 4x each points
speciesDT$dfs_m_ <- jitter(as.numeric(speciesDT$dfs_m_), amount = 2)
speciesDT$depth_m_ <- jitter(as.numeric(speciesDT$depth_m_), amount = 2)


# Create jittered duplicates
df_jittered <- df[rep(1:nrow(df), each = 3), ]  # Triplicate each row
df_jittered$y <- jitter(df_jittered$y, amount = 1)  # Jitter y

# Create a column to differentiate original vs jittered points
df$original <- TRUE
df_jittered$original <- FALSE

# Combine original and jittered data
df_combined <- rbind(df, df_jittered)

# View the final dataset
print(df_combined)

speciesDT %>% 
  select(dfs_m_, depth_m_) %>% 
  head(12)



```


